---
title: "WA_WIP"
format: html
editor: visual
---

## WA WIP

## Load libraries

```{r}
library(terra)
library(tidyterra)
library(stringr)
library(dplyr)
library(randomForest)
library(caret)
library(ggplot2)
library(lme4)
set.seed(11)
```

Load all the HUCs -8

```{r}
wa_shp <- vect("data/wa_shp_diss.gpkg")
wa_hucs <- vect("data/Hydrography/WA_HUC8.gpkg")
wa_hucs$name
```

```{r}
datapath <- paste0(getwd(), "/data/huc_data")
datapath
```

```{r}
list_rasts <- as.list(list.files(path = datapath, 
                                 full.names = TRUE, 
                                 pattern = "Skykomish",
                                 include.dirs = FALSE))
list_rasts

list_pts <- as.list(list.files(path = paste0(datapath, "/points"), 
                                 full.names = TRUE, 
                                 pattern = "Skykomish",
                                 include.dirs = FALSE))
list_pts
```



```{r}
terr_rast <- (list_rasts[grep('terr', (list_rasts))])[[1]]
(terr_rast)

deparse(substitute(terr_rast))
```


```{r}
wip_rf_func <- function(hucs, path) {
    for(i in 1:length(hucs)){
        nameClean <- str_replace_all(paste0(values(hucs[i])), "[^[:alnum:]]", "")
        rasts <- as.list(list.files(path = path, 
                                     full.names = TRUE, 
                                     pattern = nameClean,
                                     include.dirs = FALSE))
    
        pts <- as.list(list.files(path = paste0(path, "/points"), 
                                     full.names = TRUE, 
                                     pattern = nameClean,
                                     include.dirs = FALSE))
      
        training_pts <- vect(pts[[1]])
        terr_rast <- rast((rasts[grep('terr', (rasts))])[[1]])
        spec_rast <- rast((rasts[grep('spec', (rasts))])[[1]])
        clim_rast <- rast((rasts[grep('clim', (rasts))])[[1]])
        
        pts_ext <-  training_pts |>
                terra::extract(x = terr_rast, bind = T) |> 
                terra::extract(x = spec_rast, bind = T) |> 
                terra::extract(x = clim_rast, bind = T) |> 
                as.data.frame() |> 
                dplyr::mutate(class = as.factor(class)) |>
                na.exclude() |>
                write.csv(paste0(path, "/RF_Models/", 
                                 nameClean, "rf_df", ".csv"))
        
        pts_ext <- read.csv(paste0(path, "/RF_Models/", 
                                 nameClean, "rf_df", ".csv"),
                            stringsAsFactors = T) |>
                    select(-X)

        train.index <- as.vector(sample(c(1:nrow(pts_ext)), 0.7*nrow(pts_ext), replace=F))
        train <- pts_ext[train.index, ]
        test <- pts_ext[-train.index, ]

        rf_model <- randomForest(as.factor(class) ~ ., mtry = 10, 
                         sampsize = nrow(train[train$class == "WET",]),
                         replace = TRUE, #weights = wetwt, 
                         nodesize =1,
                         ntree = 1000, na.action = na.omit,
                         importance = TRUE, data = train)
        
        test_predict <- predict(rf_model, newdata = test, type = "response") 
        cm <- caret::confusionMatrix(test_predict, as.factor(test$class))
        
        cmtocsv <- data.frame(cbind(t(cm$overall),t(cm$byClass)))
        write.csv(cmtocsv,file=paste0(path, "/RF_Models/",
                                          nameClean, "rf_model_testCM", ".csv"))
        
        save(rf_model, file = paste0(path, "/RF_Models/",
                                          nameClean, "rf_model", ".RData"))
        
        
    }
}
```

# Run for the entire WA state
```{r}
wip_rf_func(wa_hucs, datapath)
```

# Test a sample
```{r}
test <- get(load("data/huc_data/RF_Models/WillapaBayrf_model.RData"))
test$importance
```



# Visualize the accuracy results for each model
```{r}
list_testacc <- as.list(list.files(path = paste0(datapath, "/RF_Models"), 
                                 full.names = TRUE, 
                                 pattern = "CM.csv",
                                 include.dirs = FALSE))

list_testacc
empty <- list()
for(i in 1:length(list_testacc)){
    csv <- read.csv(list_testacc[[i]])
    csv$X <- substr(list_testacc[[i]], 51, 60)
    empty[[i]] <- csv
}

testacc <- bind_rows(empty)
testacc
```


Now we can try to generate a prediction in one of the watersheds
To do this we need a raster stack of the covariates in the same resolution and projection. Preferably in the resolution of the DEM used to make terrain metrics

```{r}
wa_hucs$name
```


```{r}
wa_hucs$name[wa_hucs$name == "Hoh-Quillayute"]
pred_rasts <- as.list(list.files(path = datapath, 
                                     full.names = TRUE, 
                                     #pattern = str_replace_all(paste0(values("Hoh")), "[^[:alnum:]]", ""),
                                     pattern = gsub(" ", "", paste0("Hoh")),
                                     include.dirs = FALSE))
pred_rasts
```


```{r}
pred_terr <- rast((pred_rasts[grep('terr', (pred_rasts))])[[1]])
pred_spec <- rast((pred_rasts[grep('spec', (pred_rasts))])[[1]]) |> resample(pred_terr)
pred_clim <- rast((pred_rasts[grep('clim', (pred_rasts))])[[1]]) |> resample(pred_terr)


pred_terr
pred_spec
```


```{r}
pred_stack <- c(pred_terr, pred_spec, pred_clim)
names(pred_stack)
```


```{r}
list_models <- as.list(list.files(path = paste0(datapath, "/RF_Models"), 
                                 full.names = TRUE, 
                                 pattern = "HohQuillayuterf_model",
                                 include.dirs = FALSE))
#list_models
read.csv(list_models[[1]])
pred_model <- get(load(list_models[[2]]))
vip::vip(pred_model, num_features = 15)

```

```{r}
pred_wip <- predict(pred_stack, pred_model, type = "prob", 
                    filename = "data/huc_data/WIPs/HohQuillayute_WIP.tif", 
                    overwrite = TRUE)
```     

```{r}
plot(pred_wip[[2]], ext = ext(vect("data/huc_data/points/HohQuillayute_pts.gpkg")))
```  

# Clip Hoh area

```{r}
hoh_cc <- vect("data/HOH_POLYGON_711.gpkg") |> terra::project("EPSG:3740")

hoh_30m <- crop(pred_wip, hoh_cc, mask = TRUE)
hoh_30m

ggplot() + 
    geom_spatraster(data = hoh_30m[[2]])
```  
#Putting together an iterative prediction mapping function

```{r}
wip_map_func <- function(hucs, path) {
    for(i in 1:length(hucs)){
        nameClean <- str_replace_all(paste0(values(hucs[i])), "[^[:alnum:]]", "")
        rasts <- as.list(list.files(path = path, 
                                     full.names = TRUE, 
                                     pattern = nameClean,
                                     include.dirs = FALSE))
    
        pts <- as.list(list.files(path = paste0(path, "/points"), 
                                     full.names = TRUE, 
                                     pattern = nameClean,
                                     include.dirs = FALSE))
        
        mod <- get(load(as.list(list.files(path = paste0(path, "/RF_Models"), 
                                     full.names = TRUE, 
                                     pattern = paste0(nameClean, "rf_model.RData"),
                                     include.dirs = FALSE))[[1]]))
      
        training_pts <- vect(pts[[1]])
        terr_rast <- rast((rasts[grep('terr', (rasts))])[[1]])
        spec_rast <- rast((rasts[grep('spec', (rasts))])[[1]]) |> resample(terr_rast)
        clim_rast <- rast((rasts[grep('clim', (rasts))])[[1]]) |> resample(terr_rast)

        pred_stack <- c(terr_rast, spec_rast, clim_rast)

        pred_wip <- predict(pred_stack, mod, type = "prob", 
                    filename = paste0(path, "/WIPs/", nameClean, "_WIP.tif"), 
                    overwrite = TRUE)

        }
    }
```         

```{r}
wip_map_func(wa_hucs, datapath)
```         

```{r}
list_wips <- as.list(list.files(path = paste0(datapath, "/WIPs"), 
                                 full.names = TRUE, 
                                 pattern = "_WIP.tif",
                                 include.dirs = FALSE))
list_wips_r <- lapply(list_wips, rast)
sprc_wips_r <- terra::sprc(list_wips_r)
```  

```{r}
wa_wip <- terra::mosaic(sprc_wips_r, 
                        fun = "mean", 
                        filename = "data/WIP_Maps/WA_WIP.tif")
                        
```       

```{r}
r<- rast('./data/WIP_Maps/WA_WIP.tif')
plot(r)
```  