---
title: "WIP-Prabability-Hoh 10m"
format: html
editor: visual
---

## Quarto doc for Hoh area -

Development of Base model for Hoh

## Load the pkgs

```{r}
library(terra)
library(sf)
library(spatialEco)
library(lidR)
library(tidyterra)
library(dplyr)
#library(MultiscaleDTM)
library(lubridate)
library(tmap)
library(randomForest)
library(caret)
library(forcats)
```

# Load the training points

```{r}
pts <- vect("~/Documents/Qgis/teal/Hoh_data/Training_and_Validation_Data/Hoh_labelled_sample_points_using NWI_prlimmodel_07182022.shp") %>% terra::project("EPSG:3740")

terra::gdal(lib="all")
```

# Visualize the training sample

```{r}
plot(pts )
pts
```

# Use the below only if loading the individual rasters and stacking - Topo

```{r}
# list_files <- list.files(path = "~/Documents/Qgis/teal/Hoh_data/TopoIndices", full.names = TRUE, pattern = ".tif$", include.dirs = FALSE)
# list_files
# 
# # Initialize an empty vector to store filenames
# file_names <- c()
# 
# # Load rasters and check extents
# for (file in list_files) {
#   r <- rast(file)
#   print(file)
#   print(ext(r))
#   file_name <- tools::file_path_sans_ext(basename(file))
#   file_names <- append(file_names, file_name)
# }
# terr_stack<- rast(list_files) %>% terra::project("EPSG:3740")
# 
# writeRaster(terr_stack, "~/Documents/Qgis/teal/Hoh_data/stacked/TopoHoh-stacked.tif",  gdal=c("COMPRESS=LZW"),overwrite=TRUE)
# #terr_stack <- terra::rast( "~/Documents/Qgis/teal/Hoh_data/stacked/TopoHoh-stacked.tif")
# 
# names(terr_stack) <- file_names
```

# Hydro

# Use the below only if loading the individual rasters and stacking - Hydro

#SpatExtent : 385487.33385947, 436919.33385947, 5270571.3625977, 5305423.3625977 (xmin, xmax, ymin, ymax)

```{r}
list_files <- list.files(path = "~/Documents/Qgis/teal/Hoh_data/HydroIndices", full.names = TRUE, pattern = ".tif$", include.dirs = FALSE)
list_files

# Initialize an empty vector to store filenames
file_names <- c()

# Load rasters and check extents
for (file in list_files) {
  r <- rast(file)
  print(file)
  print(ext(r))
  file_name <- tools::file_path_sans_ext(basename(file))
  file_names <- append(file_names, file_name)
}
hydro_stack<- rast(list_files[2:3]) %>% terra::project("EPSG:3740")

writeRaster(hydro_stack, "~/Documents/Qgis/teal/Hoh_data/stacked/HydroHoh-stacked.tif",   gdal=c("COMPRESS=LZW","TILED=YES", "BLOCKXSIZE=256", "BLOCKYSIZE=256"),overwrite=TRUE)
#terr_stack <- terra::rast( "~/Documents/Qgis/teal/Hoh_data/stacked/TopoHoh-stacked.tif")

names(hydro_stack) <- file_names[2:3]
```

# get Spectral

```{r}
list_files <- list.files(path = "~/Documents/Qgis/teal/Hoh_data/VegIndices", full.names = TRUE, pattern = ".tif$", include.dirs = FALSE)
list_files

# Initialize an empty vector to store filenames
file_names <- c()

# Load rasters and check extents
for (file in list_files) {
  r <- rast(file)
  print(file)
  print(ext(r))
  file_name <- tools::file_path_sans_ext(basename(file))
  file_names <- append(file_names, file_name)
}
spectral_stack<- rast(list_files[7:9]) %>% terra::project("EPSG:3740")

writeRaster(spectral_stack, "~/Documents/Qgis/teal/Hoh_data/stacked/SpectralHoh-stacked.tif",   gdal=c("COMPRESS=LZW","TILED=YES", "BLOCKXSIZE=256", "BLOCKYSIZE=256"),overwrite=TRUE)
#terr_stack <- terra::rast( "~/Documents/Qgis/teal/Hoh_data/stacked/TopoHoh-stacked.tif")

names(spectral_stack) <- file_names[7:9]
```

# Extract points from

```{r}
terr_stack <- terra::rast( "~/Documents/Qgis/teal/Hoh_data/stacked/TopoHoh-stacked.tif")
hydro_stack <- terra::rast( "~/Documents/Qgis/teal/Hoh_data/stacked/HydroHoh-stacked.tif")
spectral_stack <- terra::rast( "~/Documents/Qgis/teal/Hoh_data/stacked/SpectralHoh-stacked.tif")

spectral_stack_prj <- terra::resample(spectral_stack,hydro_stack)
all_stack <- c(terr_stack,hydro_stack,spectral_stack_prj)

pts_extract <- pts |> terra::extract(x = all_stack, bind = TRUE, xy = TRUE) 
```

# Drop the unwanted columns

```{r}
pts_exdf <- as.data.frame(pts_extract) |> mutate(class = as.factor(Class)) |> 
    select(-c(x,y,Class,cell,H_NWI__,Comments,Ez_Access,Type,Moved,Analyst,FLAG,Analyst2,PrevClass,Field,Class_simp,Source,AJS_CLASS,HGM,MH_Class,MH_HGM,Uncertaint)) %>% filter(class %in% c('UPL','WET')) %>% arrange(class) %>%  droplevels() %>%  as.data.frame()

levels(pts_exdf$class) 
```

# Split to train and test

```{r}
# Validation Set 
train.index <- as.vector(sample(c(1:nrow(pts_exdf)), 0.7*nrow(pts_exdf), replace=F))

train <- pts_exdf[train.index, ]

test <- pts_exdf[-train.index, ]

set.seed(11)
```

# Train the model

```{r}
rf_model <- randomForest((class) ~ ., mtry = 12, 
#sampsize = nrow(train[train$class == "WET",]),
                         replace = TRUE, #weights = wetwt, 
                         nodesize =1,
                         ntree = 1000, na.action = na.omit,
                         importance = TRUE, data = train)
plot(rf_model)

```

# Validate on the holdout

```{r}
rf.test <- predict(rf_model, newdata = test, type = "prob")

# Plot the predicted raster
#plot(pred_wip_prob, main = "Predicted WIP-Probability")
```

# Predict on the Hoh, prediction takes longer for the entire Hoh

```{r}
pred_wip_topohydrospectral_prob <- predict(all_stack, rf_model, fun = predict, type = "prob")
# Save the predicted raster
writeRaster(pred_wip_topohydrospectral_prob, "~/Documents/Qgis/teal/Hoh_data/stacked/TopoHydroSpectralHoh-predicted-wip-prob.tif", gdal=c("COMPRESS=LZW"),overwrite = TRUE)

# Plot the predicted raster
plot(pred_wip_topohydrospectral_prob, main = "Predicted WIP-Hydro+Topo-Spectral-Probability")

```

#Apply a Focal Operation to Smooth the Raster

#smoothing by applying a focal filter. The focal() function in terra

```{r}
# Define the size of the moving window (e.g., 3x3)
window_size <- 3  # 3x3 window

# Apply a mean filter
smoothed_pred_wip_topohydrospectral_prob <- focal(pred_wip_topohydrospectral_prob, w = matrix(1, nrow = window_size, ncol = window_size), fun = mean, na.rm = TRUE)

# Save the smoothed raster
writeRaster(smoothed_pred_wip_topohydrospectral_prob,  "~/Documents/Qgis/teal/Hoh_data/stacked/TopoHydroSpectralHoh-predicted-wip-smoothed-prbability.tif", gdal=c("COMPRESS=LZW"),overwrite = TRUE)

# Plot the predicted raster
plot(smoothed_pred_wip_topohydrospectral_prob, main = "Predicted WIP- Smoothed-Hydro+TopoSpectral")
```

```{r}
# Set up the plotting area to have 1 row and 2 columns
par(mfrow = c(1, 2))

# Plot the first raster with a color palette
plot(pred_wip_topohydrospectral_prob, main = "WIP - Hydro Topo Spectral", col = terrain.colors(10))

# Plot the second raster with a color palette and custom zlim (for WIP Probability)
#plot(smoothed_raster, main = "Raster 2", col = heat.colors(10), zlim = c(0, 100))
plot(smoothed_pred_wip_topohydrospectral_prob, main = "WIP Smoothed - Hydro Topo Spectral", col = terrain.colors(10))

# Reset the plotting layout to default
par(mfrow = c(1, 1))
```

# check whether

-   Usually Larger windows produce smoother outputs but may also blur significant features.
-   Edge effects ? when applying focal operations - as pixels near the borders of the raster may be less smooth due to fewer neighboring pixels.
